---
title: "Stat 202, Spring 2021, Lab 12"
author:
- Christina Choi  
- gawonc
- Collaborators [list any collaborators here]
date: "not due / nothing submitted"
output:
  html_document:
    code_folding: show
    theme: cosmo
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

```{r, echo=FALSE}
###########################
# STYLE EDITS: IGNORE THIS
###########################

knitr::opts_chunk$set(message = FALSE) # include this if you don't want markdown to knit messages
knitr::opts_chunk$set(warning = FALSE) # include this if you don't want markdown to knit warnings
knitr::opts_chunk$set(echo = TRUE) # set echo=FALSE to hide code from html output
```


```{r}

library("knitr")
library("rmarkdown")
library("dplyr")
library("readr")

```


# Question 1:  "Lab 12 instructions"

Before looking at this file, you need to read the "lab 12 instructions" posted on Canvas.  

In particular, note that **your shimmer accounts will be removed soon after the end of the semester** (to free up server space).  The posted lab 12 instructions explain how to save your documents from shimmer, and get you started on stand-alone R Studio.

This markdown file is intended as practice for you to run in stand-alone R Studio.



# Question 2:  Not due / nothing submitted

This file is only provided as practice for those students who want to try running a markdown file in stand-alone R Studio.

Lab 12 is not graded, and there is nothing to submit.



# Question 3:  Loading Packages

The posted "lab 12 instructions" explain how to download R and R Studio on your own machine.

If you have downloaded R and R Studio for the first time, there will be no packages loaded.  When you attempt to execute a code chunk, R Studio will alert you that required packages need to be installed; you should install them as indicated.  *NOTE*:  Often, when packages need to be installed, you will get an alert that associated packages also need to be installed; you should install them as indicated.  Installing packages for the first time may take a while depending on your connection and bandwidth.



# Question 4: Loading Data

There are a variety of ways to load data in R.  One simple way is to call an online data file.  Run the following chunk to load the data file from lab 9:
*Note:*  If you are running this in newly-installed stand-alone R Studio for the first time, you will get alerts about required packages; you should click to install whatever packages and associated packages are indicated.


```{r warning = FALSE, comment = FALSE}

heart_disease <- readr::read_csv("http://stat.cmu.edu/~gordonw/heart_disease.csv")

```



# Question 5:  Excecuting an operation

Once you have successfully loaded the dataset from the chunk above, execute the following to produce a histogram of the cholesterol (`chol`) values:

```{r}

hist(heart_disease$chol,
     xlab="cholesterol",
     ylab="frequency",
     main="cholesterol of 303 patients")

```


# Question 5:  Knitting to HTML

After successfully executing the previous chunks, knit the file to html (select knit->Knit to HTML)


Congratulations!  You have successfully gotten started with your own R Studio.



**Lab 12 is not graded, and there is nothing to submit.**


****

[end of lab 12]



